---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(forcats)
library(scales)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}

TidyTueday_data_2020_09_22 <- tt_load('2020-09-22')
expeditions <- TidyTueday_data_2020_09_22$expeditions

top15_peak <- expeditions %>% 
  mutate(peak_name = fct_lump(peak_name, n = 15)) %>% 
  filter(peak_name != "Other") %>% 
  group_by(peak_name, season) %>%
  summarise(expeditions_count = n(), .groups = "drop") %>%
  mutate(peak_name = fct_reorder(peak_name, expeditions_count, .fun = sum))

ggplot(top15_peak, aes(x = expeditions_count, y = peak_name, fill = season)) +
  geom_bar(stat = "identity", position = "stack") + 
  scale_fill_viridis_d() +
  theme_light() +
  theme(legend.position = "bottom") +
  labs(
    title = "The 15 most popular peaks stacked by season of expedition",
    x = "Number of expeditions",
    y = NULL,
    fill = "Season"
  )


```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
TidyTueday_data_2019_02_19 <- tt_load('2019-02-19')
phd_by_field <- TidyTueday_data_2019_02_19$phd_by_field

phd_by_year_field <- phd_by_field %>%
  group_by(year, broad_field) %>%
  summarize(phd_count = sum(n_phds, na.rm = T))

ggplot(phd_by_year_field, aes(x = year, y = phd_count, color = broad_field, group = broad_field)) +
  geom_line(size = 1.2) +
  scale_x_continuous(breaks = pretty_breaks()) +
  scale_y_continuous(labels = comma_format()) +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() +
  labs(title = "Number of awarded Ph.D.-s in the US by Year",
       x = NULL,
       y = NULL,
       color = "Broad field")

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
TidyTueday_data_2019_11_05 <- tt_load('2019-11-05')
commute <- TidyTueday_data_2019_11_05$commute

commute_by_state <- commute %>%
  group_by(state_abb, mode) %>%
  mutate(total_commutes = sum(n, na.rm = T)) %>% 
  select(state_abb, state_region, mode, total_commutes) %>% 
  distinct() %>% 
  pivot_wider(names_from = mode, values_from = total_commutes)

ggplot(commute_by_state, aes(x = Walk, y = Bike, color = state_region)) +
  geom_point(size = 2) +
  scale_x_log10(labels = comma_format()) +
  scale_y_log10(labels = comma_format(), limits = c(20, NA)) +
  geom_text(aes(label = state_abb),  color = "black", na.rm = TRUE) +
  #geom_text_repel(aes(label = state_abb),
   #               color = "black",
    #              max.overlaps = )
  theme_light() +
  labs(title = "Title number of people walking vs. biking to work in each USA state",
       x = "Number of ppl walking to work (log N)",
       y = "Number of ppl biking to work (log N)",
       color = "State region")
  
```
